---
title: "Project 1_submission2"
author: "Luwei Zeng"
date: "2024-08-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
Build a function to create the plots you made for Presentation 
1,incorporating any feedback you received on your submission. 
Your functions should take the following input: 
(1) the name of the data frame,
(2) a list of 1 or more gene names, 
(3) 1 continuous covariate, and 
(4) two categorical covariates 

Select 2 additional genes (for a total of 3 genes) to look at and implement a loop to generate your figures using the function you created (10 pts)

```{r}
library(tidyverse)
```
One gene:A1BG;
Three genes:A1BG,A4GNT,and A1CF;
One continuous covariate:ferritin(ng/ml);
Two categorical covariates:sex, icu status.
```{r}
#Load data
setwd("/Users/hekaiwei/Desktop/R class/project")
gene <- read.csv(file = "QBS103_GSE157103_genes.csv",row.names=1)
metadata <- read.csv(file = "QBS103_GSE157103_series_matrix.csv", row.names = 1)

#convert to dataframe and switch column and row for multiple genes
selected_genes<-c("A1BG","A4GNT","A2M") #select "A1BG,A4GNT,A2M"
gene_data<-gene[selected_genes,] 
gene_data <- as.data.frame(t(gene_data)) #convert to dataframe and switch column and row

#combine selected genes expression and metadata
data_combined<-merge(gene_data,metadata,by = "row.names")
colnames(data_combined)[1] <- "ID" #set the 1st column name as "ID"
```


```{r}
#Create a function from the datasource that include gene expression, one continuous covariate,
#and two categorical covariates
plot_created<-function(data_source,gene_names,
                       continuous_covariate,
                       categorical_covariate1,categorical_covariate2){

  for (gene_name in gene_names){
      #Ensure that continuous covariate is numeric
      data_source[[continuous_covariate]] <- as.numeric(data_source[[continuous_covariate]]) 
    
  
      #Histogram for for gene expression
      plot1 <- ggplot(data_source, aes_string(x = gene_name)) +
      #Add histogram data with specific bin width, fill color, and border color
      geom_histogram(binwidth=0.04, fill="darkgreen",color="grey50")+
      labs(title = paste("Histogram of", gene_name, "Gene Expression"),
        x=paste(gene_name,"Gene Expression"),
        y="Frequency")+
      theme_classic(base_family = 'Courier',base_size = 10)

      print(plot1)

      #Scatterplot for gene expression and one continuous covariate
      plot2<-ggplot(data_source, aes_string(x =continuous_covariate, 
                                            y = gene_name,
                                            color=continuous_covariate)) +
      #Add points to the plot, setting the color of the points to dark green
      geom_point() + 
      geom_smooth(method=lm)+ #add trendline
      geom_rug(sides="bl")+ #visualize the density of the data
      labs(title=paste("Scatterplot of",gene_name,"Gene Expression vs", continuous_covariate),
        x = continuous_covariate,
        y = paste(gene_name,"Gene Expression"))+
      ylim(0,1)+
      scale_x_continuous(limits = c(0, 6000), breaks = seq(0, 6000, by = 500)) +
      #add color gradient
      scale_color_gradient(low = "green", high = "red") +
      theme_classic(base_family = 'Courier',base_size = 10)
      
      print(plot2)
  
      #Boxplot for gene expression by categorical covariates
      plot3<-ggplot(data_source, aes_string(x = categorical_covariate1, 
                                            y = gene_name, fill = categorical_covariate2)) +
      #Add boxplot
      geom_boxplot()+
      labs(title = paste("Boxplot of",gene_name,"Gene Expression by",
                         categorical_covariate1, "and",categorical_covariate2),
        x = categorical_covariate1,
        y = paste(gene_name,"Gene Expression"),
        fill = categorical_covariate2)+
      theme_classic(base_family = 'Courier',base_size = 10)+
      #Customize the theme to position the legend at the top of the plot
      theme(legend.position = 'top')

      print(plot3)
  }
}
```

```{r}
#Generate plots for A1BG by calling the function
plot_created(data_source = data_combined,
             gene_names = 'A1BG',
             continuous_covariate = "ferritin.ng.ml.",
             categorical_covariate1 = 'sex',
             categorical_covariate2 = 'icu_status')
```

```{r}
#Generate plots for A1BG,A4GNT,and A1CF by calling the function 
plot_created(data_source = data_combined,
             gene_names = selected_genes,
             continuous_covariate = "ferritin.ng.ml.",
             categorical_covariate1 = 'sex',
             categorical_covariate2 = 'icu_status')
```
